### 2017-01-26 manuscript.rmd
code review - added comments, consolidated and documented functions, added to readme

The KMRR study considered several publicly available datasets. Here we consider their conclusions in the context of one of these studies, Brawand *et al.* 2011 [@Brawand:2011du]. This dataset of gene expression for six organs across 10 species (nine mammals and one bird) is the best sampled in their analyses. Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression. It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue). It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded. This includes information about which tissue expression is specific to. For example, if one gene has expression specific to the brain and another expression specific to the kidney, both would have a Tau of 1.

@@

	# Set biology starting parameters, including files

	gene_tree_name = "Compara.75.protein.nhx.emf"

	kmrr_directory = "kmrr/"

	expression_file_names = c(
		"ChickenBrawandTScomparisonTable_9_6chPC.txt", 
		"ChimpBrawandTScomparisonTable_9_6cmPC.txt", 
		"GorillaBrawandTScomparisonTable_9_6gPC.txt", 
		"HumBrawandTScomparisonTable_9_6hPC.txt", 
		"MacacaBrawandTScomparisonTable_9_6mcPC.txt", 
		"MusBrawandTScomparisonTable_9_6mPC.txt", 
		"OpossumBrawandTScomparisonTable_9_6oPC.txt", 
		"PlatypusBrawandTScomparisonTable_9_6pPC.txt"
	)

	expression_file_names %<>% paste( kmrr_directory, ., sep="" )

	gene_tree_name %<>% paste( kmrr_directory, ., sep="" )

@@

	# These are from the KMRR code
	calibration_times = data.frame(
		age.min =
			c( 20, 92, 167, 8, 42, 74, 296, 535, 104, 937, 29, 722, 441, 65, 15, 1215, 
				414, 371, 162, 25, 74, 77, 86 ), 
		clade = 
			c("Hominoidea", "Euarchontoglires", "Mammalia", "Homininae", 
					"Simiiformes", "Primates", "Amniota", "Vertebrata", "Eutheria", 
					"Bilateria", "Catarrhini", "Chordata", "Euteleostomi", 
					"Haplorrhini", "Hominidae", "Opisthokonta", "Sarcopterygii", 
					"Tetrapoda", "Theria", "Murinae", "Sciurognathi", "Rodentia", "Glires"
			),
		stringsAsFactors=FALSE
	)

@@

	# Returns a logical vector indicating which rows in nhx_tags are for tips
	is.tip.nhx = function( nhx ) {
		is.tip = rep( FALSE, nrow( nhx@data ) )
		is.tip[ 1:length( nhx@phylo$tip.label ) ] = TRUE
		is.tip
	}

	# put node ages in @data
	store_node_age = function( nhx ) {

@@

	# Make tree ultrametric, calibrating speciation nodes to consistent times
	calibrate_tree = function ( nhx, calibration_times ) {

@@

@@

	expression = 
	lapply(
		expression_file_names, 
		read.table, 
		stringsAsFactors=FALSE, 
		header=TRUE
	) %>%
	bind_rows()

@@

```{r load_trees, echo=FALSE, cache=TRUE, message=F, warning=F}
	lines = readLines( gene_tree_name )

	# Only consider the lines with trees
	lines = lines[ grepl( "^\\(", lines, perl = TRUE ) ]

	gene_trees = mclapply(
		lines,
		function( x ){
			tree_tc = textConnection( x )
			tree = treeio::read.nhx( tree_tc )
			close( tree_tc )
			tree@data$label = c( tree@phylo$tip.label, tree@phylo$node.label )
			return( tree )
		}, mc.cores=cores
	)
```

@@

	gene_trees_annotated = mclapply(
		gene_trees,
		function( tree ){
			tree@data %<>% left_join( expression, by = c( "G" = "Ensembl.Gene.ID" ) )
			return( tree )
		}, 
		mc.cores=cores
	)

@@

	drop_empty_tips = function( nhx ) {
		to_drop = which( is.na( nhx@data[ 1:length( nhx@phylo$tip.label ), ]$Tau ) )
		remaining = length( nhx@phylo$tip.label ) - length( to_drop )

		if( remaining >= min_genes_with_expression ){
			pruned = treeio::drop.tip( nhx, to_drop )
			return( pruned )
		}
		else{
			return( NA )
		}
	}

	gene_trees_sub = mclapply( gene_trees_annotated, drop_empty_tips, mc.cores=cores )
	gene_trees_sub = gene_trees_sub[ ! is.na( gene_trees_sub ) ]

	gene_trees_calibrated = 
		mclapply( 
			gene_trees_sub,
			calibrate_tree, 
			calibration_times=calibration_times,
			mc.cores=cores
		)

	gene_trees_calibrated = gene_trees_calibrated[ ! is.na( gene_trees_calibrated ) ]

	gene_trees_calibrated = 
		mclapply( 
			gene_trees_calibrated, 
			store_node_age, 
			mc.cores=cores 
		)

@@

	pic.nhx = function( nhx ) {

		p = pic( nhx@data$Tau[ is.tip.nhx( nhx ) ], nhx@phylo, var.contrasts=TRUE )

		nhx@data$pic = c( rep( NA, length( nhx@phylo$tip.label ) ), p[ ,1 ] )
		nhx@data$var_exp = c( rep( NA, length( nhx@phylo$tip.label ) ), p[ ,2 ] )

	gene_trees_pic = mclapply( gene_trees_calibrated, pic.nhx, mc.cores=cores )

	all_nodes = 
		lapply( 
			gene_trees_pic, 
			function( nhx ){
				tags = nhx@data
				tags$gene = digest( nhx )
				tags %<>% select( -( B ) ) # B has inconsistent types, remove it
				return( tags )
			}
		) %>%
		bind_rows()

	all_nodes$Event = NA

	all_nodes$Event[ all_nodes$D == "Y" ] = "Duplication"
	all_nodes$Event[ all_nodes$D == "N" ] = "Speciation"

@@

	nodes_contrast = 
		all_nodes %>%
		filter( ! is.na( pic ) ) %>%
		mutate( pic = abs( pic ) ) %>%
		filter( ! is.na( D ) )

	nodes_contrast %>%
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
		geom_freqpoly( binwidth=.001, position="identity" ) +
		xlab( "Phylogenetic independent \ncontrasts (PIC)" ) +
		ylab( "Density" ) +
		xlim( 0, 0.075 ) + 
		theme_classic()

	wilcox_test_result = 
		wilcox.test( 
			nodes_contrast %>% filter( D=="Y" ) %>% .$pic, 
			nodes_contrast %>% filter( D=="N" ) %>% .$pic, 
			alternative="greater" 
		)

@@

We considered only the data from Brawand *et al.* 2011 [@Brawand:2011du]. KMRR examined multiple subsets of the data. We left in sex chromosome genes and testes expression data, as this corresponded to the analyses that provided the strongest support for the ortholog conjecture and therefore the most conservative reconsideration of it.

@@

>>>>

The KMRR study considered several publicly available datasets. Here we consider their conclusions in the context of one of these studies, Brawand *et al.* 2011 [@Brawand:2011du]. This dataset of gene expression for six organs across ten species (nine mammals and one bird), eight of which were analyzed by KMRR, is the best sampled in their analyses. Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression. It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue). It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded. This includes information about which tissue expression is specific to. For example, if one gene has expression specific to the brain and another expression specific to the kidney, both would have a Tau of 1.

@@

	# Name of the compara gene trees file
	gene_tree_name = "Compara.75.protein.nhx.emf"

	# Files priduced by KMRR Rscript.R with Tau values and other expression statistics
	expression_file_names = c(
		"ChickenBrawandTScomparisonTable_9_6chPC.txt", 
		"ChimpBrawandTScomparisonTable_9_6cmPC.txt", 
		"GorillaBrawandTScomparisonTable_9_6gPC.txt", 
		"HumBrawandTScomparisonTable_9_6hPC.txt", 
		"MacacaBrawandTScomparisonTable_9_6mcPC.txt", 
		"MusBrawandTScomparisonTable_9_6mPC.txt", 
		"OpossumBrawandTScomparisonTable_9_6oPC.txt", 
		"PlatypusBrawandTScomparisonTable_9_6pPC.txt"
	)

	kmrr_directory = "kmrr/"
	expression_file_names %<>% paste( kmrr_directory, ., sep="" )
	gene_tree_name %<>% paste( kmrr_directory, ., sep="" )

@@

	# These clade dates are from the KMRR code, they got them from http://timetree.org/
	calibration_times = data.frame(
		age.min =
			c( 20, 92, 167, 8, 42, 74, 296, 535, 104, 937, 29, 722, 441, 65, 15, 1215, 
				414, 371, 162, 25, 74, 77, 86 ), 
		clade = 
			c("Hominoidea", "Euarchontoglires", "Mammalia", "Homininae", 
					"Simiiformes", "Primates", "Amniota", "Vertebrata", "Eutheria", 
					"Bilateria", "Catarrhini", "Chordata", "Euteleostomi", 
					"Haplorrhini", "Hominidae", "Opisthokonta", "Sarcopterygii", 
					"Tetrapoda", "Theria", "Murinae", "Sciurognathi", "Rodentia", "Glires"
			),
		stringsAsFactors=FALSE
	)

@@

	#' Get a boolean vector corresponding to all tips and internal nodes of 
	#' a tree, with value TRUE for tips and FALSE for internal nodes
	#' 
	#' @param nhx A phylogenetic tree as a treeio::treedata object
	#' @return A boolean vector
	#' @export
	is.tip.nhx = function( nhx ) {
		is.tip = rep( FALSE, nrow( nhx@data ) )
		is.tip[ 1:length( nhx@phylo$tip.label ) ] = TRUE
		is.tip
	}

	#' Record the distance from a tip for each node in a tree in 
	#' the data slot
	#' 
	#' @param nhx A phylogenetic tree as a  treeio::treedata
	#' @return A treeio::treedata object, with an additional node_age column
	#' in the data slot, or the original object if it was not of class treedata
	#' @export
	store_node_age = function( nhx ) {

@@

	#' Adjust branch lengths of a phylogenetic tree to make it ultrametric and 
	#' to time calibrate the speciation nodes to fixed values
	#' 
	#' @param nhx A phylogenetic tree as a treeio::treedata object, with speciation 
	#' nodes annotated with a value of "N" in @data$D
	#' @param calibration_times A dataframe with two columns: age.min is the age 
	#' of a clade, and clade is the name of the clade
	#' @return A treeio::treedata object if successfully calibrated
	#' @export
	calibrate_tree = function ( nhx, calibration_times ) {

@@

	#' Drop tips without expression data from a tree 
	#' 
	#' @param nhx A phylogenetic tree as a treeio::treedata object, with a 
	#' column @data$Tau containing expression data
	#' @param min_genes_with_expression The minimum number of tips with 
	#' expression data for a tree to be retained
	#' @return A treeio::treedata object, or NA if less than 
	#' min_genes_with_expression tips would have been retained
	#' @export
	drop_empty_tips = function( nhx, min_genes_with_expression ) {

		# Identify the tips without expression data
		to_drop = which( is.na( nhx@data[ 1:length( nhx@phylo$tip.label ), ]$Tau ) )

		# Return the pruned tree if it meets sampling criteria, otherwise return NA
		remaining = length( nhx@phylo$tip.label ) - length( to_drop )
		if( remaining >= min_genes_with_expression ){
			pruned = treeio::drop.tip( nhx, to_drop )
			return( pruned )
		}
		else{
			return( NA )
		}
	}

	#' Calculate phylogenetic independent contrasts (PIC) with the tree and 
	#' character data in a treeio::treedata object
	#' 
	#' @param nhx A phylogenetic tree as a treeio::treedata object, with a 
	#' column @data$Tau containing expression data
	#' @return A treeio::treedata object, with new @data columns pic and var_exp
	#' @export
	pic.nhx = function( nhx ) {

		# Calculate the contrasts
		p = pic( nhx@data$Tau[ is.tip.nhx( nhx ) ], nhx@phylo, var.contrasts=TRUE )

		# Add the results back to the @data slot, padding the rows that correspond to 
		# tips with NA
		nhx@data$pic = c( rep( NA, length( nhx@phylo$tip.label ) ), p[ ,1 ] )
		nhx@data$var_exp = c( rep( NA, length( nhx@phylo$tip.label ) ), p[ ,2 ] )

@@

	# Read the expression data for each species and combine into a single tibble
	expression = 
	lapply(
		expression_file_names, 
		read.table, 
		stringsAsFactors=FALSE, 
		header=TRUE
	) %>%
	bind_rows()

@@

```{r load_trees, echo=FALSE, cache=TRUE, message=F, warning=F}

	# Read in the tree file as a vector of strings, one per line
	lines = readLines( gene_tree_name )

	# Only retain strings that are trees
	lines = lines[ grepl( "^\\(", lines, perl = TRUE ) ]

	# Parse each string to a treeio::treedata object, adding a label column 
	# for node names
	gene_trees = mclapply(
		lines,
		function( x ){
			tree_tc = textConnection( x )
			tree = treeio::read.nhx( tree_tc )
			close( tree_tc )
			tree@data$label = c( tree@phylo$tip.label, tree@phylo$node.label )
			return( tree )
		}, mc.cores=cores
	)
```

@@

	# Annotate each tree by joining the corresponding expression data to the 
	# @data object. This maps the expression data to the tree tips
	gene_trees_annotated = mclapply(
		gene_trees,
		function( tree ){
			tree@data %<>% left_join( expression, by = c( "G" = "Ensembl.Gene.ID" ) )
			return( tree )
		}, 
		mc.cores=cores
	)

@@

	# Remove tips from the trees that do not have expression data
	gene_trees_sub = mclapply( 
		gene_trees_annotated, 
		drop_empty_tips, 
		min_genes_with_expression=min_genes_with_expression,
		mc.cores=cores 
	)

	# Remove trees with less than min_genes_with_expression tips with expression data
	gene_trees_sub = gene_trees_sub[ ! is.na( gene_trees_sub ) ]

	# Make the trees ultrametric and calibrate the speciation nodes to 
	# specified times
	gene_trees_calibrated = 
		mclapply( 
			gene_trees_sub,
			calibrate_tree, 
			calibration_times=calibration_times,
			mc.cores=cores
		)

	# Remove trees that could not be successfully calibrated
	gene_trees_calibrated = gene_trees_calibrated[ ! is.na( gene_trees_calibrated ) ]

	# Parse the node ages from the internal @phylo object and store them with the 
	# corresponding rows in the @data object
	gene_trees_calibrated = 
		mclapply( 
			gene_trees_calibrated, 
			store_node_age, 
			mc.cores=cores 
		)

@@

	# Calculate independent contrasts for Tau on each tree, storing the results 
	# back into the @data slot of the tree objects
	gene_trees_pic = mclapply( gene_trees_calibrated, pic.nhx, mc.cores=cores )

	# Combine the @data slots across all trees. These have labels, node annotations,
	# expression data, contrast results, etc...
	all_nodes = 
		lapply( 
			gene_trees_pic, 
			function( nhx ){
				tags = nhx@data
				tags$gene = digest( nhx ) # Creates a  hash that is unique to each gene tree
				tags %<>% select( -( B ) ) # B has inconsistent types, remove it
				return( tags )
			}
		) %>%
		bind_rows()

	# Make a new human readable column for plots that records speciation/ duplication
	all_nodes$Event = NA
	all_nodes$Event[ all_nodes$D == "Y" ] = "Duplication"
	all_nodes$Event[ all_nodes$D == "N" ] = "Speciation"

@@

	# Create a subset that corresponds to nodes with contrasts. This excludes tips.
	# Also take the absolute value of the contrast, since we only consider magnitude.
	nodes_contrast = 
		all_nodes %>%
		filter( ! is.na( pic ) ) %>%
		mutate( pic = abs( pic ) ) %>%
		filter( ! is.na( D ) )

	# Plot the contrasts for duplication events and speciation events
	nodes_contrast %>%
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
		geom_freqpoly( binwidth=.001, position="identity" ) +
		xlab( "Phylogenetic independent \ncontrasts (PIC)" ) +
		ylab( "Density" ) +
		xlim( 0, 0.075 ) + 
		theme_classic()

	# Check if the mean rank of duplication pics is greater than speciation pics
	wilcox_test_result = 
		wilcox.test( 
			nodes_contrast %>% filter( D=="Y" ) %>% .$pic, 
			nodes_contrast %>% filter( D=="N" ) %>% .$pic, 
			alternative="greater" 
		)

@@

We considered only the data from Brawand *et al.* 2011 [@Brawand:2011du] for the eight taxa included in KMRR. We left in sex chromosome genes and testes expression data, as this corresponded to the analyses that provided the strongest support for the ortholog conjecture and therefore the most conservative reconsideration of it.

@@

	# Plot results of the ascertainment bias analyses
